---
title: "Demographic Predictors of Reform UK Vote Share in the 2024 General Election"
author: "Anton Robb"
output:
  pdf_document:
    toc: true
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 7, fig.height = 5)
```

# Introduction

The 2024 UK general election saw Reform UK emerge as a significant electoral force,
with support varying widely across the country's 650 parliamentary constituencies.
This project asks a simple question: which demographic and socioeconomic
characteristics of a constituency best predict how strongly it voted for Reform UK?

To answer it, I combine constituency-level election results with 2021 census
demographics (age structure, ethnic composition, educational attainment,
benefit-claimant rates and median wages) and constituency geography. I build a
multiple linear regression model of Reform's vote share, and pay particular
attention to a statistical subtlety that turns out to matter a great deal here:
how the apparent effect of a variable can change, and even reverse, once other
correlated variables are accounted for.

All data are public, sourced from the House of Commons Library and the 2021
census. Data preparation, modelling and evaluation are carried out entirely in R.

# Preparing the data

The analysis draws on several separate files, one per data domain, each keyed by a
constituency identifier code. I load each, keep a focused set of columns to avoid
redundancy and overfitting, and merge them into a single constituency-level frame.

```{r load-merge}
library(dplyr)

elec  <- read.csv("election2024.csv")
age   <- read.csv("age.csv")
claim <- read.csv("claimants.csv")
eth   <- read.csv("ethnicity.csv")
qual  <- read.csv("qualifications.csv")
wage  <- read.csv("wages.csv")

# Retain a small, non-redundant set of explanatory variables per domain
age   <- age[,   c("ONSConstID", "pct18.24", "pct65.up")]
claim <- claim[, c("ONSConstID", "claimant_rate")]
eth   <- eth[,   c("ONSConstID", "white", "asian", "black")]
qual  <- qual[,  c("ONSConstID", "Qual4_pct", "Qual5_pct")]
wage  <- wage[,  c("ONSConstID", "MedianWage")]

elec$Region <- factor(elec$RegionName)

df <- elec |>
  merge(age,   by = "ONSConstID") |>
  merge(claim, by = "ONSConstID") |>
  merge(eth,   by = "ONSConstID") |>
  merge(qual,  by = "ONSConstID") |>
  merge(wage,  by = "ONSConstID")
```

The response variable is Reform's share of the valid vote in each constituency.
I restrict the analysis to constituencies where Reform actually stood a candidate,
since a zero vote elsewhere reflects absence rather than lack of support.

```{r response}
df$vote_share <- df$RUK / df$Valid.votes
df <- df[df$RUK > 0, ]
```

I selected the explanatory variables to capture distinct demographic and
socioeconomic dimensions while avoiding near-duplicates: two age bands (younger and
older adults), three broad ethnic groups, two educational-attainment measures (at
opposite ends of the qualification scale), the benefit-claimant rate, median wage,
and Region as a categorical geographic factor. The aim was breadth of genuinely
different signals rather than many collinear measures of the same underlying trait.

# Exploring the response

```{r response-dist}
library(ggplot2)

ggplot(df, aes(vote_share)) +
  geom_histogram(binwidth = 0.025, fill = "turquoise4", colour = "white") +
  labs(title = "Distribution of Reform UK vote share across constituencies",
       x = "Vote share", y = "Count") +
  theme_minimal()

summary(df$vote_share)
```

Reform's vote share is unimodal and mildly right-skewed, concentrated between
roughly 5% and 25% per constituency, with a small number of constituencies where
Reform performed exceptionally strongly forming a right tail.

# Univariate associations

As a first pass, I examine each continuous variable's association with vote share
individually: the slope of a simple linear fit, the Pearson correlation, and its
p-value. This gives a marginal picture, before any variables are considered
together.

```{r univariate}
cont_vars <- c("pct18.24", "pct65.up", "white", "asian", "black",
               "Qual4_pct", "Qual5_pct", "claimant_rate", "MedianWage")

uni <- lapply(cont_vars, function(v) {
  ct <- cor.test(df$vote_share, df[[v]])
  data.frame(variable = v,
             slope = coef(lm(df$vote_share ~ df[[v]]))[2],
             correlation = unname(ct$estimate),
             p_value = ct$p.value)
}) |> do.call(what = rbind)

uni[order(uni$p_value), ]
```

The strongest single association is with the proportion of residents holding no
formal qualifications, which is strongly *negatively* correlated with Reform
support. Several other variables (age structure, ethnic composition, median wage)
also show clear marginal associations.

# Regional variation

Region is a categorical variable, so I test its association with vote share using a
one-way analysis of variance, and inspect the regional means.

```{r region}
aov_region <- aov(vote_share ~ Region, data = df)
summary(aov_region)

tapply(df$vote_share, df$Region, mean) |> sort()
```

Region is very strongly associated with Reform support: average vote share differs
substantially across the country, with the highest support in parts of the North
and Midlands and the lowest in London and Scotland. Geography clearly carries real
signal.

# A combined ranking of marginal associations

Bringing the continuous and categorical results together, and correcting the
p-values for multiple comparisons, gives a ranked view of which variables are most
strongly associated with vote share when each is considered on its own.

```{r combined-ranking}
p_region <- summary(aov_region)[[1]]["Region", "Pr(>F)"]

ranking <- rbind(
  data.frame(variable = cont_vars, p_value = uni$p_value),
  data.frame(variable = "Region", p_value = p_region)
)
ranking$p_adj <- p.adjust(ranking$p_value, method = "bonferroni")
ranking[order(ranking$p_adj), ]
```

# Multiple regression

The univariate view is only a starting point, because the explanatory variables are
themselves correlated with one another. To isolate each variable's *independent*
contribution, I fit a multiple linear regression including all explanatory
variables simultaneously.

```{r multi-model}
multi <- lm(vote_share ~ pct18.24 + pct65.up + white + asian + black +
              Qual4_pct + Qual5_pct + claimant_rate + MedianWage + Region,
            data = df)
summary(multi)
```

## Model diagnostics

```{r diagnostics, fig.height=6}
par(mfrow = c(2, 2))
plot(multi)
par(mfrow = c(1, 1))
```

The diagnostic plots show broadly well-behaved residuals: approximately normal with
only mild departures in the tails, and roughly constant variance. A small number of
high-leverage constituencies are present but none dominate the fit. (Any single
constituency with both extreme residual and high leverage would be worth excluding
and refitting; the diagnostics here do not indicate a case severe enough to distort
the conclusions.)

## Model summary statistics

```{r model-stats}
s <- summary(multi)
e <- residuals(multi); h <- hatvalues(multi)
data.frame(
  df_model      = length(coef(multi)) - 1,
  df_error      = df.residual(multi),
  R_squared     = s$r.squared,
  adj_R_squared = s$adj.r.squared,
  residual_SE   = s$sigma,
  AIC           = AIC(multi),
  PRESS         = sum((e / (1 - h))^2)
)
```

The model explains a large majority of the variation in Reform vote share across
constituencies, indicating that constituency demographics account for much of the
geographic pattern in support.

# Univariate versus multivariate: where the story changes

The most interesting statistical feature of this analysis is how variables behave
differently in isolation versus in the full model. Comparing each variable's
covariate-adjusted p-value from the multiple regression against its marginal
p-value reveals three distinct patterns.

```{r compare}
coef_tab <- s$coefficients
p_cont_multi <- coef_tab[cont_vars, "Pr(>|t|)"]
p_region_multi <- anova(multi)["Region", "Pr(>F)"]

compare <- data.frame(
  variable = c(cont_vars, "Region"),
  p_univariate = ranking$p_value[match(c(cont_vars, "Region"), ranking$variable)],
  p_multivariate = c(p_cont_multi, p_region_multi)
)
compare[order(compare$p_multivariate), ]
```

Three things stand out:

*Variables that stay strong.* Educational attainment, age structure and Region
remain highly significant in the full model, indicating genuine independent
associations with Reform support that are not merely proxies for something else.

*A suppressed effect that emerges.* The benefit-claimant rate is weak or
insignificant on its own but becomes clearly significant once education, age and
region are held constant. This is a suppression effect: the raw association is
masked by other variables, and only controlling for them reveals the underlying
relationship between economic deprivation and vote share.

*Associations that dissolve, and one that reverses.* Some variables that looked
important in isolation, such as ethnic composition, lose significance in the full
model, implying their marginal associations were largely explained by correlation
with education and region. Most strikingly, the older-population share flips sign:
positively associated with Reform support on its own, but *negatively* associated
once other variables are controlled for. Older populations tend to co-occur with
other constituency characteristics (particularly around education and region) that
themselves drive support, and once those are accounted for, the residual
partial association with age runs the other way.

# Interpretation

```{r illustrate, fig.height=4}
par(mfrow = c(1, 2))
plot(df$Qual5_pct, df$vote_share, pch = 16, col = "turquoise4",
     xlab = "Proportion with no qualifications", ylab = "Reform vote share",
     main = "Vote share vs education")
abline(lm(vote_share ~ Qual5_pct, data = df), col = "firebrick", lwd = 2)

plot(df$pct65.up, df$vote_share, pch = 16, col = "turquoise4",
     xlab = "Proportion aged 65+", ylab = "Reform vote share",
     main = "Vote share vs older population")
abline(lm(vote_share ~ pct65.up, data = df), col = "firebrick", lwd = 2)
par(mfrow = c(1, 1))
```

Educational composition is the single strongest structural predictor of Reform's
constituency-level performance, and its effect is robust: it survives controlling
for age, ethnicity, income and region. The age result is the clearest cautionary
tale of the analysis. Viewed alone, older constituencies appear more supportive of
Reform, but this bivariate impression is misleading; once the confounding influence
of education and region is removed, the partial relationship reverses. It is a
concrete demonstration of why simple pairwise relationships can mislead when
explanatory variables are correlated, and why multiple regression, interpreted
carefully, gives a more trustworthy picture of what independently drives an outcome.

# Notes

Data: 2024 general election results and 2021 census demographics, sourced from the
House of Commons Library (public). This project began as a piece of university
coursework and has been rewritten as an independent, self-directed analysis. All
processing is reproducible from the original public data files in R.